Bird beak's elaboration and innovation

### 2022-11-09 Analysis/07-Orthogonality_and_statistics.Rmd
Update 07-Orthogonality_and_statistics.Rmd

---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This section contains the code for reproducing the analyses. To get the details, refer to [`dispRity` vignette](link_to_projection_vignette).
This script take around 6 minutes to compile.

@@

```{r, eval = FALSE, echo = FALSE}
## Load the data for all birds
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace_allbirds <- shapespace_allbirds_lvl_superorder_order[[1]]$space
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

@@

```{r, eval = FALSE, echo = FALSE}
## Creating the nested levels subgroups
make.nested.subset <- function(group, data, group.data) {
    ## Get the subsets within the group
    level2_groups <- as.character(unique(group.data$level2[group.data$level1 == group]))
    if(any(empty <- level2_groups == "")) {
        level2_groups <- level2_groups[-which(empty)]
    }
    if(length(level2_groups) < 1) {
        return(NULL)
    }
    ## Create a subset
    return(get.subsets(data, subsets = c(level2_groups, group)))
}

@@

```{r, eval = FALSE, echo = FALSE}
source("../Functions/stats.utilities.R")

@@

```{r, echo = FALSE, eval = TRUE}
## Quick load
source("../Functions/stats.utilities.R")
library(dispRity)
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
results <- shapespace_allbirds_lvl_superorder_order_results_list
rm(shapespace_allbirds_lvl_superorder_order_results_list)

## Distribution of elaboration and innovation at the species level
elaboration_species      <- get.disparity(get.subsets(results$super_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]]
innovation_species       <- get.disparity(get.subsets(results$super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]]

## Distribution of elaboration and innovation at the clade level
scale.group <- function(X) return(X/max(X))
elaboration_super_orders <- lapply(get.disparity(results$super_results_between$position, concatenate = FALSE), scale.group)
innovation_super_orders  <- lapply(get.disparity(results$super_results_between$distance, concatenate = FALSE), scale.group)
elaboration_orders       <- lapply(get.disparity(results$order_results_between$position, concatenate = FALSE), scale.group)
innovation_orders        <- lapply(get.disparity(results$order_results_between$distance, concatenate = FALSE), scale.group)
## The objects above are the ones that contain all the elaboration and innovation values by clade level

## These bits are for the Bhatt coeff
elaboration_group <- rbind(do.call(rbind, elaboration_super_orders), do.call(rbind, elaboration_orders))
innovation_group <- rbind(do.call(rbind, innovation_super_orders), do.call(rbind, innovation_orders))

@@

pdf("../Manuscript/Figures/relative_EI.pdf", height = 6, width = 12)
par(mfrow = c(1,2))
plot.density(species = c(elaboration_species),
             group   = c(unlist(elaboration_super_orders), unlist(elaboration_orders)),
             xlab    = "relative elaboration", scale.x = TRUE,
             ylab    = "density",
             main    = "A - Total posterior elaboration per clade\n(scaled) and per species")
legend("topright", lty = 1, legend = c("species", "clades"), col = c("blue", "orange"))
text(x = 0.7, y = 0.5, paste0("Posterior area under the curve overlap\n(Bhattacharyya coefficient): ", round(bhatt.coeff(elaboration_species_auc, elaboration_group_auc), 3)))
plot.density(species = c(innovation_species),
             group   = c(unlist(innovation_super_orders), unlist(innovation_orders)),
             xlab    = "relative innovation", scale.x = TRUE,
             ylab    = "density",
             main    = "B - Total posterior innovation per clade\n(scaled) and per species")
text(x = 0.7, y = 0.5, paste0("Posterior area under the curve overlap\n(Bhattacharyya coefficient): ", round(bhatt.coeff(innovation_species_auc, innovation_group_auc), 3)))
dev.off()
```

>>>>

---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme and Gavin Thomas"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This section contains the code for reproducing the analyses. To get the details, refer to [`dispRity` vignette](link_to_projection_vignette).
This script take around 35 minutes to compile.

@@

```{r, eval = TRUE, echo = FALSE}
## Load the data for all birds
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace_allbirds <- shapespace_allbirds_lvl_superorder_order[[1]]$space
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

@@

```{r, eval = TRUE, echo = FALSE}
## Creating the nested levels subgroups
make.nested.subset <- function(group, data, group.data) {
    ## Get the subsets within the group
    level2_groups <- as.character(unique(group.data$level2[group.data$level1 == group]))
    if(any(empty <- level2_groups == "")) {
        level2_groups <- level2_groups[-which(empty)]
    }
    if(length(level2_groups) < 1) {
        return(NULL)
    }
    ## Create a subset
    return(get.subsets(data, subsets = c(level2_groups, group)))
}

@@

```{r, eval = TRUE, echo = FALSE}
source("../Functions/stats.utilities.R")

@@

```{r, echo = FALSE, eval = TRUE}
source("../Functions/stats.utilities.R")
library(dispRity)
library(ggplot2)
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
results <- shapespace_allbirds_lvl_superorder_order_results_list
rm(shapespace_allbirds_lvl_superorder_order_results_list)

## Distribution of elaboration and innovation at the species level
elaboration_species <- get.disparity(get.subsets(results$super_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]]
innovation_species  <- get.disparity(get.subsets(results$super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]]
elaboration_species <- elaboration_species/max(elaboration_species)
innovation_species  <- innovation_species/max(innovation_species)

## Distribution of elaboration and innovation at the clade level
scale.group <- function(X) return(X/max(X))
elaboration_super_orders <- lapply(get.disparity(results$super_results_between$position, concatenate = FALSE), scale.group)
innovation_super_orders  <- lapply(get.disparity(results$super_results_between$distance, concatenate = FALSE), scale.group)
elaboration_orders       <- lapply(get.disparity(results$order_results_between$position, concatenate = FALSE), scale.group)
innovation_orders        <- lapply(get.disparity(results$order_results_between$distance, concatenate = FALSE), scale.group)

## The objects above are the ones that contain all the elaboration and innovation values by clade level
## These bits are for the Bhatt coeff
elaboration_group <- rbind(do.call(rbind, elaboration_super_orders), do.call(rbind, elaboration_orders))
innovation_group <- rbind(do.call(rbind, innovation_super_orders), do.call(rbind, innovation_orders))

@@

library(tidyverse)
library(patchwork)

fake <- data.frame(elaboration = c(elaboration_species, c(unlist(elaboration_super_orders), unlist(elaboration_orders))), innovation = c(innovation_species, c(unlist(innovation_super_orders), unlist(innovation_orders))), Level = c(rep("Species", 34992000), rep("Clades", 140000)))

elab <- 
  ggplot(fake, aes(x = elaboration, colour=Level, fill = Level, ..scaled..)) +
  geom_density(alpha = 0.5) +
  theme_classic(base_size = 14) +
  scale_colour_manual(values = c(rgb(0,0,0,0), rgb(0,0,0,0))) +
  scale_fill_manual(values = c("goldenrod", "#1E65A7")) +
  xlab("Relative elaboration") +
  ylab("Density") +
  theme(legend.position = "none") +
  labs(tags="a")

innov <-
  ggplot(fake, aes(x = innovation, colour = Level, fill = Level, ..scaled..)) +
  geom_density(alpha = 0.5) +
  theme_classic(base_size = 14) +
  scale_colour_manual(values = c(rgb(0,0,0,0), rgb(0,0,0,0))) +
  scale_fill_manual(values = c("goldenrod", "#1E65A7")) +
  xlab("Relative innovation") + 
  theme(legend.position = c(0.8, 0.8)) +
  labs(tags="b")

remove_y <- 
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

p <- elab + innov + remove_y

ggsave("relative_EI_gg_style.pdf", plot=p, height=6, width=12)
```
